---
title: "How to Conduct a Clone Censor-Weight Survival Analysis using survivalCCW"
author: Matthew Secrest
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Conduct a Clone Censor-Weight Survival Analysis using survivalCCW}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Clone Censor Weighting

This lightweight package describes how to conduct clone-censor weighting (CCW) to address the problem of immortal time bias in survival analysis. This vignette will walk through the applied tutorial published by [Maringe et al 2020](https://academic.oup.com/ije/article/49/5/1719/5835351). Refer to [Hernan and Robins 2016](https://doi.org/10.1093/aje/kwv254) and [Hernan et al 2016](https://doi.org/10.1016/j.jclinepi.2016.04.014) for more technical details.

# Context

CCW is useful in the presence of immortal person-time bias in observational studies. For instance, when comparing surgery recipients vs non-recipients in non-small cell lung cancer (NSCLC), the surgery group will have a longer survival time than the non-surgery group because the non-surgery group includes patients who died before they could receive surgery. This is a form of immortal time bias.

The CCW toy dataset published by Maringe uses this exact setting as the motivating example. Let's explore the dataset, which comes with `survivalCCW`.

```{r}
library(survivalCCW)
head(toy_df)
```

Column descriptions can be found with `?toy_df`:

- `id`: patient identifier
- `fup_obs`: observed follow-up time (time to death or 1 year if censored alive)
- `death`: observed event of interest (all-cause death) 1: dead, 0: alive
- `timetosurgery`: time to surgery (NA if no surgery)
- `age`: age at diagnosis
- `sex`: patient's sex
- `perf`: performance status at diagnosis
- `stage`: stage at diagnosis
- `deprivation`: deprivation score
- `charlson`: Charlson's comorbidity index
- `emergency`: route to diagnosis

Note that this package addresses situations in which the covariates are all defined at baseline.

# Create clones

The first step is to create the clones. This can be done for any time-to-event outcome using the `survivalCCW` function `create_clones`. For `create_clones` to work, we need to pass a one-row-per-patient `data.frame` with the following columns:

- The id variable (in this case, `id`)
- The traditional outcome variable which denotes censorship (0) or event (1) (in this case, `death`). Note that additional values are not yet permitted.
- The time to the first event (in this case, `fup_obs`)
- The exposure variable, with exposure defined **at any time prior to censorship/event** (in this case, `surgery`). Must be (0) or (1),
- The time to exposure variable (in this case, `timetosurgery`)
- The clinical eligibility window (in this case, we'll do 6 months)

All other columns will be propogated for each patient. Let's see what this looks like in practice.

```{r}

# Create clones
clones <- create_clones(
  df = toy_df,
  id = 'id',
  event = 'death',
  time_to_event = 'fup_obs',
  exposure = 'surgery',
  time_to_exposure = 'timetosurgery',
  ced_window = 365.25/2
)

head(clones)

```


Note that this object is just a `data.frame` with an additional custom class which future functions will evaluate:
```{r}
class(clones)
```


# Cast to long format
Now we simply need to cast the data to long format. The `survivalCCW` function `cast_to_long` will do this for us. 
No additional arguments are needed (the `clones` object is an artifact that allows you to better see and understand the method):

```{r}
clones_long <- cast_clones_to_long(clones)

head(clones_long)

```


# Bootstrap